The importance of reproducibility in COVID-19 research: the case of population density and the spread of the pandemic

### 2021-05-19 R0-Density-Reanalysis/R0-Density-Reanalysis.tex
First draft of introduction and background

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Population density and basic reproductive number of
COVID-19 across United States counties
revisited} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Antonio Paez\textsuperscript{\getAff{McMaster
University}}\textsuperscript{*}\\
\bigskip
\textbf{\getAff{McMaster University}}School of Earth, Environment and
Society, 1280 Main St West, Hamilton, Ontario L8S 4K1 Canada\\
\bigskip
* Corresponding author: paezha@mcmaster.ca\\
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget
porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut
dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id
velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus
finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium
diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut
ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit
pulvinar, volutpat risus.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step.
% Author Summary not valid for PLOS ONE submissions.
\section*{Author summary}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget
porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut
dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id
velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus
finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium
diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut
ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit
pulvinar, volutpat risus.

@@

% Use "Eq" instead of "Equation" for equation citations.
\emph{Text based on plos sample manuscript, see
\url{https://journals.plos.org/ploscompbiol/s/latex}}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget
porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut
dui eu ante cursus gravida non sed sem. Phasellus mauris velit, dapibus
finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium
diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut
ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit
pulvinar, volutpat risus.

A list

\begin{itemize}
\tightlist
\item
  Item 1
\item
  Item 2
\end{itemize}

Here are two sample references: {[}1,2{]}.

>>>>

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{The importance of reproducibility in COVID-19 research:
the case of population density and the spread of the
pandemic} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Antonio Paez\textsuperscript{\getAff{McMaster
University}}\textsuperscript{*}\\
\bigskip
\textbf{\getAff{McMaster University}}School of Earth, Environment and
Society, 1280 Main St West, Hamilton, Ontario L8S 4K1 Canada\\
\bigskip
* Corresponding author: paezha@mcmaster.ca\\
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
The emergence of the novel SARS-CoV-2 coronavirus and the global
COVID-19 pandemic has led to explosive growth in scientific research.
Given the high stakes of the situation, it is essential that scientific
activites, on which good policy depends, are as transparent and
reproducible as possible. Reproducibility is key for the efficient
operation of the self-correction mechanisms of science, which work to
weed out errors and refine our understanding of social and physical
phenomena. In this paper, the importance of reproducibility is
illustrated for the case of the association between population density
and the the spread of SARS-CoV-2. Transparency and openness means that
the same problem can, with relatively modest efforts, be examined by
independent researchers who can verify findings, and bring to bear
different perspectives, approaches, and methods---sometimes with
consequantial changes in the conclusions, as the empirical example in
this paper shows.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step.
% Author Summary not valid for PLOS ONE submissions.

@@

% Use "Eq" instead of "Equation" for equation citations.
\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

The emergence of the novel SARS-CoV-2 coronavirus in 2019, and the
global pandemic that followed in its wake, led to an explosive growth of
research. According to Fraser et al. {[}1{]}, over 125,000
COVID-19-related papers were released in the first ten months from the
first confirmed case of the disease. Of these, more than 30,000 were
shared in pre-print servers, the use of which also exploded in the past
year {[}2--4{]}.

Given the heavy human and economic cost of the pandemic, there has been
a natural tension in the scientific community between the need to
publish research results quickly and the imperative to maintain
consistently high quality standards in scientific reporting; indeed, a
call for maintaining the standards in published research has even called
this deluge of publications a ``carnage of substandard research''
{[}5{]}. Part of the challenge of maintaining quality standards in
published research is that, despite an abundance of recommendations and
guidelines {[}6--9{]}, in practice reproducibility has remained a lofty
but somewhat aspirational goal {[}10,11{]}. As reported in the
literature, only a woefully small proportion of published research was
actually reproducible before the pandemic {[}12,13{]}, a situation that
does not appear to have changed substantially since {[}14,15{]}.

The push for open data and software, along with more strenuous efforts
towards open, reproducible research, is simply a continuation of
long-standing scientific practices of independent verification. Despite
the (at times disproportionate) attention that high profile scandals in
science tend to elicit in the media, science as a collective endeavor is
remarkable for being a self-correcting enterprise, one with built-in
mechanisms and incentives to weed out erroneous ideas. Over the long
term, facts tend to prevail in science. At stake is the shorter-term
impacts that research may have in other spheres of economic and social
life. The case of economists Reinhart and Rogoff comes to mind: by the
time the inaccuracies in their research were uncovered {[}see 16{]},
their claims about debt and economic growth had already been seized by
policy-makers on both sides of the Atlantic to justify austerity
policies in the aftermath of the Great Recession of
2007-2009\footnote{Nobel Prize in Economics Paul Krugman noted that
  ``Reinhart--Rogoff may have had more immediate influence on public
  debate than any previous paper in the history of economics''
  \url{https://www.nybooks.com/articles/2013/06/06/how-case-austerity-has-crumbled/?pagination=false}}.
As later research has demonstrated, those policies cast a long shadow,
and their sequels continued to be felt for years {[}17{]}.

In the context of COVID-19, a topic that has grabbed the imagination of
numerous thinkers has been the prospect of life in cities after the
pandemic {[}18{]}. The fact that the worst of the pandemic was initially
felt in dense population centers such as Wuhan, Milan, Madrid, and New
York, prompted a flurry of research into the associations between
density and the spread of the pandemic. Some important questions hang on
the results of these research efforts. For example, are lower density
regions safer from the pandemic? Are de-densification policies
warranted, at least in the short term? And in the longer term, will the
risks of life in high density regions presage a flight from cities? Over
the past year, numerous papers have sought to throw light into the
underlying issue of density and the pandemic; nonetheless the results,
as will be detailed next, remain mixed. Further, to complicate matters,
precious few of these studies appear to be sufficiently open to support
independent verification.

The objective of this paper is to illustrate the importance of
reproducibility in research, particularly in the context of the flood of
COVID-19 papers. To this end, a recent study by Sy et al. {[}19{]} is
chosen as an example of reproducible research. The objective is not to
malign the analysis of these researchers, but rather to demonstrate the
value of openness to allow for independent verification and further
analysis. Open data and open code mean that an independent researcher
can, with only modest efforts, not only verify the findings reported,
but also examine the same data from a perspective which may not have
been available to the original researchers due to differences in
disciplinary perspectives, methodological traditions, and/or training,
among other possible factors. The example, which shows consequential
changes in the conclusions reached by different analyses, should serve
as a call to researchers to redouble their efforts to increase
transparency and reproducibility in research. This paper, in addition,
aims to show how data can be packaged in well-documented, shareable
units, and code can be embedded into self-contained documents suitable
for review and independent verification. The source for this paper is an
\href{http://rmarkdown.rstudio.com}{R Markdown} document which, along
with the data package, is available in a public repository\footnote{\url{https://github.com/paezha/Reproductive-Rate-and-Density-US-Reanalyzed}}.

\hypertarget{background}{%
\section{Background}\label{background}}

The concern with population density and the spread of the virus during
the COVID-19 pandemic was fueled, at least in part, by dramatic scenes
seen in real-time around the world from large urban centers such as
Wuhan, Milan, Madrid, and New York. In theory, there are good reasons to
believe that higher density may have a positive association with the
transmission of a contagious virus. It has long been known that the
potential for inter-personal contact is greater in regions with higher
density {[}see for example the research on urban fields and
time-geography 20,21,22{]}. Mathematically, models of exposure and
contagion indicate that higher densities can catalyze the transmission
of contagious diseases {[}23,24{]}. Models such as these were likely at
the root of messages, by some figures in positions of authority, that
low density regions faced lower risks from the pandemic\footnote{Governor
  Kristi Noem of South Dakota, for example, claimed that sparse
  population density allowed her state to face the pandemic without the
  need for strict policy interventions
  \url{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}}.

As Rocklöv and Sjödin {[}23{]} note, however, mathematical models of
contagion are valid at small-to-medium spatial scales (and presumably,
small temporal scales too, such as time spent in restaurants, concert
halls, cruises), and the results do not necessarily transfer to larger
spatial units and different time scales. There are good reasons for
this: while in a restaurant, one can hardly avoid being in proximity to
other customers-however, a person can choose to (or be forced to as a
matter of policy) not go to a restaurant in the first place.
Nonetheless, the idea that high density correlates with high
transmission is so intuitive that it is often taken for granted even at
larger scales {[}e.g., 25,26{]}. At larger scales, however, there exists
the possibility of behavioral adaptations, which are difficult to
capture in the mechanistic framework of differential equations {[}or can
be missing in agent-based models, 27{]}; these adaptations, in fact, can
be a key aspect of disease transmission.

A plausible behavioral adaptiation during a pandemic, especially one
broadcast as widely and intensely as COVID-19, is risk compensation.
Risk compensation is a process whereby people adjust their behavior in
response to their \emph{perception} of risk {[}28--30{]}. For example,
it is possible that people who listened to the message of leaders saying
that they were safe because of low density may not have taken adequate
precautions. People in dense places who could more directly observe the
impact of the pandemic may have become overly cautious. Both Paez et al.
{[}31{]} and Hamidi et al. {[}32{]} posit this mechanism (i.e., greater
compliance with social distancing in denser regions) to explain the
results of their analyses. The evidence available does indeed show that
there were important changes in behavior during the pandemic, at least
with respect to mobility {[}33--35{]}; furthermore, shelter in place
orders may have had greater buy-in from the public in higher density
regions {[}36{]}, and the behavior may have persisted beyond the
duration of official social-distancing policies {[}37{]}. In addition,
there is evidence that changes in mobility correlated with the
trajectory of the pandemic {[}38,39{]}. Given the potential for
behavioral adaptation, the question of density becomes more nuanced: it
is not just a matter of proximity, but also of human behavior, which is
better studied using population-level data and models.

In this respect, the literature to date remains inconclusive.

On the one hand, there are studies that report positive associations
between population density and various COVID-19-related outcomes. Bhadra
{[}40{]}, for example, reported a moderate positive correlation between
the spread of COVID-19 and population density at the district level in
India, however their analysis was bivariate and did not control for
other variables, such as income. Similarly, Kadi and Khelfaoui {[}41{]}
found a positive and significant correlation between number of cases and
population density in cities in Algeria in a series of simple regression
models (i.e., without other controls). A question in these relatively
simple analyses is whether density is not a proxy for other factors.
Other studies have included controls, such as Pequeno et al. {[}42{]}, a
team that reported a positive association between density and cumulative
counts of confirmed COVID-19 cases in state capitals in Brazil after
controlling for covariates, including income, transport connectivity,
and economic status. In a similar vein, Fielding-Miller et al. {[}43{]}
reported a positive relationship between the absolute number of COVID-19
deaths and population density (rate) in rural counties in the US. Roy
and Ghosh {[}44{]} used a battery of machine learning techniques to find
discriminatory factors, and a positive and significant association
between COVID-19 infection and death rates in US states. Wong and Li
{[}45{]} also found a positive and significant association between
population density and number of confirmed COVID-19 cases in US
counties, using both univariate and multivariate regressions with
spatial effects. Most recently, Sy et al. {[}19{]} reported that the
basic reproductive number of COVID-19 in US counties tends to increase
with population density, but at a decreasing rate at higher densities.

On the gripping hand, a number of studies report non-significant or
negative associations between population density and COVID-19 outcomes.
This includes the research of Sun et al. {[}46{]} who did not find
evidence of significant correlation between population density and
confirmed number of cases per day \emph{in conditions of lockdown} in
China. This finding echoes the results of Paez et al. {[}31{]}, who in
their study of provinces in Spain reported non-significant associations
between population density and infection rates in the early days of the
first wave of COVID-19, and negative significant associations in the
later part of the first lockdown. Similarly, {[}47{]} found zero or
negative associations between population density and infection
numbers/deaths by country. Fielding-Miller et al. {[}43{]} found a
negative relationship between COVID-19 deaths and population density
urban counties in the US. In their investigation of doubling time, White
and Hébert-Dufresne {[}48{]} identified a negative and significant
correlation between population density and doubling time in US states.
Likewise, {[}49{]} fond a small negative (and significant) association
between population density and COVID-19 morbidity in districts in
Tehran. And two of the most complete studies in the US by Hamidi et al.
{[}32,50{]} used an extensive set of controls to find negative and
significant correlations between density at the level of counties in the
US and COVID-19 cases and fatalities.

As can be seen, these studies are implemented at different scales in
different regions of the world. They also use a range of techniques,
from correlation analysis, to multivariate regression, spatial
regressions, and machine learning techniques. This is natural and to be
expected: individual researchers have only limited time and expertise.
This is why reproducibility is important. To pick an example (which will
be further elaborated in the following sections), the study of Sy et
al.~{[}{[}19{]}; hereafter SWN{]} would immediately grab the attention
of a researcher with expertise in spatial modelling. Such an expert
would likely ask some of the following questions: how were missing
counties treated? Is it possible to spatially interpolate missing
observations? What are the implications of the spatial sampling
framework used in the analysis? Was there evidence of spatial
autocorrelation in the residuals of the models? These are questions that
in most cases would not occur to a researcher who has not been exposed
to spatial statistics or spatial econometrics. Nonetheless, they are
relevant and important. Fortunately, SWN give an example of a reasonably
open, reproducible research product: their paper is accompanied by (most
of) the data and (most of) the code used in the analysis. This means
that an independent expert can, with only a moderate investment of time
and effort, replicate the results in the paper, as well as ask
additional questions.

Alas, reproducibility is not necessarily the norm.

There are various reasons why a project can fail to be reproducible. In
some cases, there might be legitimate reasons to withhold the data,
perhaps due to confidentiality and privacy reasons {[}e.g., 51{]}. But
in many other cases the data are publicly available, as has been
commonly the case with population-level COVID-19 information. Often the
provenance of the data is documented, but the data themselves are not
shared {[}25,32,40,43,50,e.g., 52,53--55{]}. As any researcher can
attest, whether a graduate student or a seasoned scientist, collecting,
organizing, and preparing data for a project can take a substantial
amount of time. Pointing to the sources of data, even when these sources
are public, is a small step towards reproducibility-but only a very
small one. Faced with the prospect of having to recreate a data set from
raw sources is probably sufficient to dissuade all but the most
dedicated (or stubborn) researcher. This is true even if part of the
data are shared {[}e.g., 45{]}. In other cases, data are shared, but the
processes to document the preparation of the data are not fully
documented {[}47,e.g., 56{]}. These processes matter, as shown by the
errors in the spreadsheets of Reinhart and Rogoff {[}16{]}, and the data
of biologist Jonathan Pruitt that led to an ``avalanche'' of paper
retractions\footnote{\url{https://doi.org/10.1038/d41586-020-00287-y}}.
Another situation is when papers share well-documented data, but fail to
provide the code used in the analysis {[}42,e.g., 57,58{]}. Making code
available only ``on demand'' {[}e.g., 59{]} is an unnecessary barrier
when most journals offer the facility to share supplemental materials
online. Then there are those papers that strive towards reproducibility,
sharing well-documented processes and data, as well as the code used in
any analyses reported {[}e.g., 31,36,48,60{]}.

\hypertarget{replicating-swn}{%
\subsection{Replicating SWN}\label{replicating-swn}}

Fit (mixed) linear models as in Sy et al.:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{table1model1 }\OtherTok{\textless{}{-}} \FunctionTok{lme}\NormalTok{(R }\SpecialCharTok{\textasciitilde{}}\NormalTok{ density\_log , }
                    \AttributeTok{random =} \SpecialCharTok{\textasciitilde{}} \DecValTok{1}\SpecialCharTok{|}\NormalTok{ state, }
                    \AttributeTok{data =}\NormalTok{ county\_geo\_clean }\SpecialCharTok{\%\textgreater{}\%}
                      \FunctionTok{filter}\NormalTok{(R }\SpecialCharTok{\textgreater{}} \DecValTok{0}\NormalTok{))}
\FunctionTok{summary}\NormalTok{(table1model1)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Linear mixed-effects model fit by REML
##   Data: county_geo_clean %>% filter(R > 0) 
##        AIC      BIC    logLik
##   2383.621 2403.808 -1187.811
## 
## Random effects:
##  Formula: ~1 | state
##         (Intercept)  Residual
## StdDev:    0.165504 0.6648325
## 
## Fixed effects:  R ~ density_log 
##                 Value  Std.Error   DF  t-value p-value
## (Intercept) 2.2740045 0.05468629 1099 41.58272       0
## density_log 0.1621127 0.01485543 1099 10.91269       0
##  Correlation: 
##             (Intr)
## density_log 0.797 
## 
## Standardized Within-Group Residuals:
##        Min         Q1        Med         Q3        Max 
## -2.3847932 -0.5972261 -0.1682299  0.3681864 16.2716371 
## 
## Number of Observations: 1151
## Number of Groups: 51
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{intervals}\NormalTok{(table1model1)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Approximate 95% confidence intervals
## 
##  Fixed effects:
##                 lower      est.     upper
## (Intercept) 2.1667032 2.2740045 2.3813058
## density_log 0.1329645 0.1621127 0.1912609
## attr(,"label")
## [1] "Fixed effects:"
## 
##  Random Effects:
##   Level: state 
##                     lower     est.     upper
## sd((Intercept)) 0.1080424 0.165504 0.2535261
## 
##  Within-group standard error:
##     lower      est.     upper 
## 0.6375478 0.6648325 0.6932849
\end{verbatim}

Linear mixed models of R0 and density + \% private transportation

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{table1model2 }\OtherTok{\textless{}{-}} \FunctionTok{lme}\NormalTok{(R }\SpecialCharTok{\textasciitilde{}}\NormalTok{ density\_log }\SpecialCharTok{+}\NormalTok{ private , }
                    \AttributeTok{random =} \SpecialCharTok{\textasciitilde{}} \DecValTok{1}\SpecialCharTok{|}\NormalTok{ state, }
                    \AttributeTok{data =}\NormalTok{ county\_geo\_clean }\SpecialCharTok{\%\textgreater{}\%}
                      \FunctionTok{filter}\NormalTok{(R }\SpecialCharTok{\textgreater{}} \DecValTok{0}\NormalTok{))}
\FunctionTok{summary}\NormalTok{(table1model2)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Linear mixed-effects model fit by REML
##   Data: county_geo_clean %>% filter(R > 0) 
##        AIC      BIC    logLik
##   2385.478 2410.707 -1187.739
## 
## Random effects:
##  Formula: ~1 | state
##         (Intercept)  Residual
## StdDev:   0.1358322 0.6649714
## 
## Fixed effects:  R ~ density_log + private 
##                 Value Std.Error   DF   t-value p-value
## (Intercept)  3.347070 0.3418910 1098  9.789874  0.0000
## density_log  0.145258 0.0156188 1098  9.300177  0.0000
## private     -0.012547 0.0039434 1098 -3.181760  0.0015
##  Correlation: 
##             (Intr) dnsty_
## density_log -0.222       
## private     -0.988  0.344
## 
## Standardized Within-Group Residuals:
##        Min         Q1        Med         Q3        Max 
## -2.7347592 -0.6005758 -0.1610880  0.3876853 16.2104980 
## 
## Number of Observations: 1151
## Number of Groups: 51
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{intervals}\NormalTok{(table1model2)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Approximate 95% confidence intervals
## 
##  Fixed effects:
##                   lower       est.        upper
## (Intercept)  2.67623619  3.3470697  4.017903160
## density_log  0.11461177  0.1452579  0.175904019
## private     -0.02028416 -0.0125468 -0.004809444
## attr(,"label")
## [1] "Fixed effects:"
## 
##  Random Effects:
##   Level: state 
##                      lower      est.    upper
## sd((Intercept)) 0.08059718 0.1358322 0.228921
## 
##  Within-group standard error:
##     lower      est.     upper 
## 0.6376445 0.6649714 0.6934695
\end{verbatim}

Linear mixed models of R0 and density + \% private transportation +
median income

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{table1model3 }\OtherTok{\textless{}{-}} \FunctionTok{lme}\NormalTok{(R }\SpecialCharTok{\textasciitilde{}}\NormalTok{ density\_log }\SpecialCharTok{+}\NormalTok{ private }\SpecialCharTok{+}\NormalTok{ hincome, }
                    \AttributeTok{random =} \SpecialCharTok{\textasciitilde{}} \DecValTok{1}\SpecialCharTok{|}\NormalTok{ state, }
                    \AttributeTok{data =}\NormalTok{ county\_geo\_clean }\SpecialCharTok{\%\textgreater{}\%}
                      \FunctionTok{filter}\NormalTok{(R }\SpecialCharTok{\textgreater{}} \DecValTok{0}\NormalTok{))}
\FunctionTok{summary}\NormalTok{(table1model3)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Linear mixed-effects model fit by REML
##   Data: county_geo_clean %>% filter(R > 0) 
##       AIC     BIC   logLik
##   2393.98 2424.25 -1190.99
## 
## Random effects:
##  Formula: ~1 | state
##         (Intercept)  Residual
## StdDev:   0.1373995 0.6651151
## 
## Fixed effects:  R ~ density_log + private + hincome 
##                 Value Std.Error   DF   t-value p-value
## (Intercept)  3.385550 0.3933810 1097  8.606289  0.0000
## density_log  0.146826 0.0171534 1097  8.559585  0.0000
## private     -0.012707 0.0040548 1097 -3.133786  0.0018
## hincome     -0.003256 0.0151105 1097 -0.215493  0.8294
##  Correlation: 
##             (Intr) dnsty_ privat
## density_log  0.027              
## private     -0.950  0.211       
## hincome     -0.493 -0.412  0.228
## 
## Standardized Within-Group Residuals:
##        Min         Q1        Med         Q3        Max 
## -2.7162603 -0.5984817 -0.1626793  0.3887931 16.2029477 
## 
## Number of Observations: 1151
## Number of Groups: 51
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{intervals}\NormalTok{(table1model3)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Approximate 95% confidence intervals
## 
##  Fixed effects:
##                   lower         est.       upper
## (Intercept)  2.61368596  3.385550103  4.15741424
## density_log  0.11316880  0.146825984  0.18048317
## private     -0.02066315 -0.012707018 -0.00475089
## hincome     -0.03290488 -0.003256195  0.02639249
## attr(,"label")
## [1] "Fixed effects:"
## 
##  Random Effects:
##   Level: state 
##                    lower      est.     upper
## sd((Intercept)) 0.081461 0.1373995 0.2317504
## 
##  Within-group standard error:
##     lower      est.     upper 
## 0.6377582 0.6651151 0.6936455
\end{verbatim}

Linear mixed models of R0 and density + \% private transportation +
median income + interaction of density and \% private transportation

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{table1model4 }\OtherTok{\textless{}{-}} \FunctionTok{lme}\NormalTok{(R }\SpecialCharTok{\textasciitilde{}}\NormalTok{ density\_log}\SpecialCharTok{*}\NormalTok{private }\SpecialCharTok{+}\NormalTok{ hincome, }
                    \AttributeTok{random =} \SpecialCharTok{\textasciitilde{}} \DecValTok{1}\SpecialCharTok{|}\NormalTok{ state, }
                    \AttributeTok{data =}\NormalTok{ county\_geo\_clean }\SpecialCharTok{\%\textgreater{}\%}
                      \FunctionTok{filter}\NormalTok{(R }\SpecialCharTok{\textgreater{}} \DecValTok{0}\NormalTok{))}
\FunctionTok{summary}\NormalTok{(table1model4)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Linear mixed-effects model fit by REML
##   Data: county_geo_clean %>% filter(R > 0) 
##       AIC      BIC   logLik
##   2407.04 2442.349 -1196.52
## 
## Random effects:
##  Formula: ~1 | state
##         (Intercept)  Residual
## StdDev:   0.1378096 0.6653589
## 
## Fixed effects:  R ~ density_log * private + hincome 
##                         Value Std.Error   DF   t-value p-value
## (Intercept)          3.415859 0.4157861 1096  8.215424  0.0000
## density_log          0.177645 0.1350641 1096  1.315266  0.1887
## private             -0.013123 0.0044568 1096 -2.944502  0.0033
## hincome             -0.002719 0.0153117 1096 -0.177607  0.8591
## density_log:private -0.000354 0.0015414 1096 -0.229894  0.8182
##  Correlation: 
##                     (Intr) dnsty_ privat hincom
## density_log          0.323                     
## private             -0.952 -0.386              
## hincome             -0.409  0.105  0.139       
## density_log:private -0.323 -0.992  0.414 -0.158
## 
## Standardized Within-Group Residuals:
##        Min         Q1        Med         Q3        Max 
## -2.7371582 -0.5968733 -0.1638499  0.3883700 16.1980628 
## 
## Number of Observations: 1151
## Number of Groups: 51
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{intervals}\NormalTok{(table1model4)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Approximate 95% confidence intervals
## 
##  Fixed effects:
##                            lower          est.        upper
## (Intercept)          2.600032192  3.4158588336  4.231685475
## density_log         -0.087368297  0.1776452292  0.442658756
## private             -0.021867960 -0.0131231030 -0.004378246
## hincome             -0.032762972 -0.0027194668  0.027324038
## density_log:private -0.003378809 -0.0003543603  0.002670088
## attr(,"label")
## [1] "Fixed effects:"
## 
##  Random Effects:
##   Level: state 
##                      lower      est.     upper
## sd((Intercept)) 0.08169505 0.1378096 0.2324679
## 
##  Within-group standard error:
##     lower      est.     upper 
## 0.6379755 0.6653589 0.6939176
\end{verbatim}

\hypertarget{some-relevant-questions}{%
\section{Some relevant questions}\label{some-relevant-questions}}

Alas, despite decades' worth of developments in the field of
geographical analysis, not all research presented to date has used
proper methods for the study of COVID-19. While the answer to the
question ``do spatial effects really matter in regression analysis'' was
definitively answered in the positive at least 40 years ago, in practice
many researchers continue to ignore the pitfalls of ignoring them. In
this paper, I present a reanalysis of the data used by Sy et al.~(2021)
to study the correlations between the basic reproductive number of
COVID-19 and population density in counties. I highlight two related
issues: non-systematic sampling in space and spatial autocorrelation.
The reanalysis is based on the use of tobit models to account for
non-systematic sampling, and spatially autoregressive tobit models to
account for spatial autocorrelation in the data generation process. The
reanalysis highlights the importance of openness and reproducibility in
COVID-19 research. Finally, the results provide a sobering example of
the risks of not using appropriate methods in the analysis of
geographical data. Furthermore,

\hypertarget{fit-tobit-version-of-models}{%
\subsection{Fit tobit version of
models}\label{fit-tobit-version-of-models}}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{table1model1 }\OtherTok{\textless{}{-}} \FunctionTok{censReg}\NormalTok{(R }\SpecialCharTok{\textasciitilde{}}\NormalTok{ density\_log, }
                        \AttributeTok{left =} \DecValTok{0}\NormalTok{,}
                    \AttributeTok{data =}\NormalTok{ county\_geo\_clean)}
\FunctionTok{summary}\NormalTok{(table1model1)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## 
## Call:
## censReg(formula = R ~ density_log, left = 0, data = county_geo_clean)
## 
## Observations:
##          Total  Left-censored     Uncensored Right-censored 
##           3218           2067           1151              0 
## 
## Coefficients:
##             Estimate Std. error t value Pr(> t)    
## (Intercept)  2.33967    0.09232   25.34  <2e-16 ***
## density_log  0.74323    0.02672   27.82  <2e-16 ***
## logSigma     0.56894    0.02396   23.74  <2e-16 ***
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## Newton-Raphson maximisation, 7 iterations
## Return code 8: successive function values within relative tolerance limit (reltol)
## Log-likelihood: -3246.5 on 3 Df
\end{verbatim}

Linear mixed models of R0 and density + \% private transportation

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{table1model2 }\OtherTok{\textless{}{-}} \FunctionTok{censReg}\NormalTok{(R }\SpecialCharTok{\textasciitilde{}}\NormalTok{ density\_log }\SpecialCharTok{+}\NormalTok{ private, }
                    \AttributeTok{data =}\NormalTok{ county\_geo\_clean)}
\FunctionTok{summary}\NormalTok{(table1model2)}
\end{Highlighting}
\end{Shaded}
